Language learning in context via recursive pragmatic reasoning

### 2013-05-31 praglearn.tex
tiny tweaks to 5.2

A stronger example of how pragmatics can create communicative biases can be
observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
In these games, pragmatic reasoning causes each agent to start out weakly biasedtowards the associations ``cheap'' $\leftrightarrow$
\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
is uniform and uninformative. A fully rational listener, therefore,
observing an uncertain speaker using this mapping, would discount it
as arising from this bias, and conclude that the speaker in fact did
not know which convention was in use. Conventional listeners,
however, believe that speakers do know which convention is in
use, and therefore tend to take their behavior as positive evidence that the
``good'' system is in use. Similarly, conventional speakers
will wager that since on average they will succed more often if
listeners are using the ``good'' system, they might as well try
it. When they succeed, they take their success as evidence that the listener was in
fact using the good system all along. As a result, dyads in this game
 end up converging onto a stable system at a rate far above
chance (Fig.~\ref{fig:horn}).

Something
fundamentally different happens in this case than in the specificity
implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
saw that our model could hear pragmatically strengthened usages
of words while recovering the underlying literal meaning. In the Horn
case, though, the pragmatic strengthening depends essentially on
ignorance about literal meaning. As the agents gather more data,
the implicature disappears, to be replaced by a belief that
``cheap'' \textit{literally} means \textsc{common}. To demonstrate
this phenomenon, we queried each agent in each dyad about how they would refer to
or interpret each object and word, \textit{if} the two objects were
equally common, which cancels the Horn implicature. As shown in
Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
of dyads both $S$ and $L$ believed in the ``good'' mapping even in
this implicature-free case, while less than 20\% believed in the
``bad'' mapping (with the rest being inconsistent). This makes the
interesting prediction that while specificity implicatures can survive
indefinitely in a language system, any Horn implicatures that arise
consistently should transition to becoming part of the literal meaning
of the words involved.

>>>>

A stronger example of how pragmatics can create communicative biases can be
observed by considering the Horn implicatures introduced in \S\ref{sec:horn-implicature}. 
In these games, pragmatic reasoning causes each agent to start out
weakly biased towards the associations ``cheap'' $\leftrightarrow$
\textsc{common}, ``expensive'' $\leftrightarrow$ \textsc{rare}; this bias arises because their prior on literal meaning
is uniform and uninformative. A fully rational listener, therefore,
observing an uncertain speaker using this mapping, would discount it
as arising from this bias, and conclude that the speaker in fact did
not know which convention was in use. Conventional listeners,
however, believe that speakers do know which convention is in
use, and therefore tend to take their behavior as positive evidence that the
``good'' system is in use. Similarly, conventional speakers
will wager that since on average they will succed more often if
listeners are using the ``good'' system, they might as well try
it. When they succeed, they take their success as evidence that the listener was in
fact using the good system all along. As a result, dyads in this game
 end up converging onto a stable system at a rate far above
chance (Fig.~\ref{fig:horn}).

Something
fundamentally different happens in this case than in the specificity
implicature case (\S\ref{sec:learning-specificity-implic}), however. There, we
saw that our model could hear pragmatically strengthened usages
of words while recovering the underlying literal meaning. In the Horn
case, though, the pragmatic strengthening depends essentially on
ignorance about literal meaning. As the agents gather more data,
the implicature disappears, to be replaced by a belief that
``cheap'' \textit{literally} means \textsc{common} (and did all along). To demonstrate
this phenomenon, we queried each agent in each dyad about how they would refer to
or interpret each object and word, \textit{if} the two objects were
equally common, which cancels the Horn implicature. As shown in
Fig.~\ref{fig:emergence-average} (right), after 30 turns, in nearly 70\%
of dyads both $S$ and $L$ believed in the ``good'' mapping even in
this implicature-free case, while less than 20\% believed in the
``bad'' mapping (with the rest being inconsistent). Our model therefore makes the
interesting prediction that while specificity implicatures can survive
indefinitely in a language system, any Horn implicatures that arise
consistently should transition to becoming part of the literal meaning
of the words involved.
